Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-09-01 paper/main.tex
caught a typo

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{newfloat}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{placeins}

@@

\begin{document}
\maketitle

\begin{abstract}
Our ongoing subjective experience reflects external sensory information from each moment, along with additional information from our past that we carry with us into that moment.  The blend of memories, knowledge, emotions, goals, and other internal perceptual and mental states that color our subjective experience provides a \textit{context} for interpreting new information and conceptually linking what is happening now with our prior experiences.  Because this contextual information is often person-specific, the subjective experience that each person encodes into their memory is often idiosyncratic, even for shared experiences and sensory perspectives.  We sought to study which aspects of a shared naturalistic experience were preserved or distorted, and how those distortions compared across individuals.  To this end, we developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences.  We model experiences as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration.  We also demonstrate how \textit{memories} may also be modeled as trajectories through the same spaces.  According to this view, encoding an experience into memory entails geometrically preserving, distorting, or transforming the \textit{shape} of the original experience's trajectory.  This translates qualitative neuropsychological questions about how we remember naturalistic experiences into quantitive geometric questions about the spatial configurations of trajectory shapes.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectories of participants' recountings of the episode nearly all captured the coarse spatial properties of the original episode's trajectory (i.e., the essential plot points), but participants differed in their memory for low-level details.  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains preserve, distort, and transform our ongoing experiences when we encode them into episodic memories.
\end{abstract}

@@

Here, we sought to examine how the temporal dynamics of a naturalistic experience were later reflected in participants' memories.  We also sought to leverage the above conceptual insights into the distinctions between an experience's essence and its low-level details to build models that explicitly quantified these distinctions.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television show \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Our framework uses topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls and hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast both naturalistic experiences and memories of those experiences as geometric \textit{trajectories} that describe how they evolve over time. Under this framework, successful remembering entails verbally traversing the content trajectory of the episode, thereby reproducing the shape (essence) of the original experience.  Our framework captures the episode's essence in the sequence of geometric coordinates for its events, and its low-level details by examining its within-event geometric properties.

@@

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall transcript, we treated each overlapping  window of (up to 10) sentences from their transcript as a document, and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  An important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite that different participants used widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This ability to match up conceptually related text that differs in specific vocabulary, detail, and length is an important benefit of projecting the episode and recalls into a shared topic space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall trajectory (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

@@

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the episode, and the events they later recalled.  We labeled each recalled event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transitioned between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging the meaningful narrative associations and references over long timescales throughout the episode.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of episode events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of episode events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision metric reflects completeness of recall.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  episode events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} Excerpts from the most precise (P17) and least precise (P6) participants' recalls of episode event 22.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.}
  \label{fig:precision-detail}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/distinctiveness_detail}
  \caption{\small \textbf{Distinctiveness metric reflects specificity of recall.} \textbf{A.} Recall distinctiveness by episode event.  Kernel density estimates for each episode event's distribution of recall distinctiveness scores, analogous to Fig.~\ref{fig:precision-detail}A.  \textbf{B.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in panel C.  Each event's text is highlighted in a different color.  \textbf{C.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 22.  Sections of recall describing each each episode event in Panel B are highlighted with the corresponding color.}
  \label{fig:distinctiveness-detail}
\end{figure}

Examining individual recalls of the same episode event can provide insights into how the above precision and distinctiveness scores may be used to characterize similarities and differences in how different people describe the same shared experience.  In Figure \ref{fig:precision-detail}, we compare recalls for the same episode event (event 22) from two participants: one with a high precision score (P17), and the other with a low precision score (P6).  From the HMM-identified event boundaries, we recovered the set of annotations describing the content of an example episode event (Fig.~\ref{fig:precision-detail}B), and divided them into different color-coded sections for each action or feature described.  We used an analogous approach to identify the set of sentences comprising the corresponding recall events for each of the two example participants.  Figure~\ref{fig:precision-detail}C shows excerpts of two participants' recall transcripts that comprised sentences between the first and last descriptions of content from the example episode event.  We then colored all words describing actions and features in the transcripts shown in Panel C according to the color-coded annotations in Panel B.  Visual comparison of these example recalls reveals that the more precise recall captures more of the episode event's content, and in greater detail.

Figure \ref{fig:distinctiveness-detail} illustrates the differences between high and low distinctiveness scores for the same event detailed in Figure~\ref{fig:precision-detail} (i.e., event 22).  Here, we have extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) matched to the example episode event (Fig.~\ref{fig:distinctiveness-detail}C).  We also extracted the annotations for the example episode event, as well as those from each other episode event whose content the example participants' single recall events described (Fig.~\ref{fig:distinctiveness-detail}B).  We assigned each episode event a unique color (Panel B) and colored each recalled phrase or sentence (Panel C) according to the episode events they best matched.  Visual inspection of Panel C reveals that the most distinctive recall's content is tightly concentrated around event 22, whereas the least distinctive recall incorporates content from a much wider range of episode events.

The preceding analyses sought to characterize how participants' recountings of individual episode events captured the low-level details of each event.  Next we sought to characterize how participants' recountings of the full episode captured its high-level essence-- i.e., the shape of the episode's trajectory through topic space.  To visualize the essence of the episode and each participant's recall trajectory~\citep{HeusEtal18a}, we projected the topic proportions matrices for the episode and recalls onto a shared two-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInEtal18}.  In this lower-dimensional space, each point represents a single episode or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}). In other words, events that are nearer to each other in this space are more semantically similar, and those that are farther apart are less so.

@@

In addition to enabling us to visualize the episode's high-level essence, describing the episode as a geometric trajectory also enables us to drill down to individual words and quantify how each word relates to the memorability of each event.  This provides another approach to examining participants' recall for low-level details beyond the precision and distinctiveness measures we defined above.  The results displayed in Figures \ref{fig:list-learning}C and \ref{fig:precision-detail}A suggest that certain events were remembered better than others.  Given this, we next asked asked whether the events were generally remembered precisely or imprecisely tended to reflect particular content.  Because our analysis framework projects the dynamic episode content and participants' recalls into a shared space, and because the dimensions of that space represent topics (which are, in turn, sets of weights over known words in the vocabulary), we are able to recover the weighted combination of words that make up any point (i.e., topic vector) in this space.  We first computed the average precision with which participants recalled each of the 30 episode events (Fig.~\ref{fig:topics}A; note that this result is analogous to a serial position curve created from our precision metric).  We then computed a weighted average of the topic vectors for each episode event, where the weights reflected how precisely each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by more precisely-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Across the full episode, content that reflected topics necessary to convey the central focus of the episode (e.g., the names of the two main characters, ``Sherlock'' and ``John,'' and the address of a major recurring location, ``221B Baker Street'') were best remembered.  An analogous analysis revealed which themes were less-precisely remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to its average precision (Fig.~\ref{fig:topics}B, red box).  The least precisely remembered episode content reflected information that was extraneous to the episode's essence, such as the proper names of relatively minor characters (e.g., ``Mike,'' ``Molly,'' and ``Lestrade'') and locations (e.g., ``St. Bartholomew's Hospital'').

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/topics}
\caption{\small \textbf{Language used in the most and least precisely remembered events.} \textbf{A.} Average precision (episode event-recall event topic vector correlation) across participants for each episode event.  Here we defined each episode event's precision for each participant as the correlation between its topic vector and the most-correlated recall event's topic vector from that participant.  Error bars denote bootstrap-derived across-participant 95\% confidence intervals.  The stars denote the three most precisely remembered events (green) and least precisely remembered events (red).  \textbf{B.} Wordles comprising the top 200 highest-weighted words reflected in the weighted-average topic vector across episode events.  Green: episode events were weighted their precision (Panel A).  Red: episode events were weighted by the inverse of their precision.  \textbf{C.}  The set of all episode and recall events is projected onto the two-dimensional space derived in Figure~\ref{fig:trajectory}.  The dots outlined in black denote episode events (dot size is proportional to each event's precision).  The dots without black outlines denote recall events from each participant.  All dots are colored using the same scheme as Figure~\ref{fig:trajectory}A.  Wordles for several example events are displayed (green: three most precisely remembered events; red: three least precisely remembered events).  Within each circular wordle, the left side displays words associated with the topic vector for the episode event, and the right side displays words associated with the (average) recall event topic vector, across all recall events matched to the given episode event.}
\label{fig:topics}
\end{figure}

A similar result emerged from assessing the topic vectors for individual episode and recall events (Fig.~\ref{fig:topics}C).  Here, for each of the three most and least precisely remembered episode events, we have constructed two wordles: one from the original episode event's topic vector (left) and a second from the average recall topic vector for that event (right).  The three most precisely remembered events (circled in green) correspond to scenes integral to the central plot-line: a mysterious figure spying on John in a phone booth; John meeting Sherlock at Baker St. to discuss the murders; and Sherlock laying a trap to catch the killer.  Meanwhile, the least precisely remembered events (circled in red) reflect scenes that comprise minor plot points: a video of singing cartoon characters that participants viewed in an introductory clip prior to the main episode; John asking Molly about Sherlock's habit of over-analyzing people; and Sherlock noticing evidence of Anderson's and Donovan's affair.

The results thus far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories.  We next carried out a series of analyses aimed at understanding which brain structures might facilitate these preservations and transformations between the external world and memory.  In the first analysis, we sought to identify brain structures that were sensitive to the dynamic unfolding of the episode's content, as characterized by its topic trajectory.  We used a searchlight procedure to identify clusters of voxels whose activity patterns displayed a proximal temporal correlation structure (as participants watched the episode) matching that of the original episode's topic proportions (Fig.~\ref{fig:brainz}A; see \textit{Methods} for additional details).  In a second analysis, we sought to identify brain structures whose responses (during episode viewing) reflected how each participant would later structure their \textit{recounting} of the episode.  We used a searchlight procedure to identify clusters of voxels whose proximal temporal correlation matrices matched that of the topic proportions matrix for each participant's recall transcript (Figs.~\ref{fig:brainz}B; see \textit{Methods} for additional details).  To ensure our searchlight procedure identified regions \textit{specifically} sensitive to the temporal structure of the episode or recalls (i.e., rather than those with a temporal autocorrelation length similar to that of the episode and recalls), we performed a phase shift-based permutation correction (see \textit{Methods} for additional details). As shown in Figure~\ref{fig:brainz}C, the episode-driven searchlight analysis revealed a distributed network of regions that may play a role in processing information relevant to the narrative structure of the episode.  Similarly, the recall-driven searchlight analysis revealed a second network of regions (Fig.~\ref{fig:brainz}D) that may facilitate a person-specific transformation of one's experience into memory.  The top ten \texttt{Neurosynth} terms~\citep{YarkEtal11} associated with each (unthresholded) map are displayed in each panel.  In identifying regions whose responses to ongoing experiences reflect how those experiences will be remembered later, this latter analysis extends classic \textit{subsequent memory analyses}~\citep[e.g.,][]{PallWagn02} to the domain of naturalistic experiences.

>>>>

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{newfloat}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{placeins}

@@

\begin{document}
\begin{titlepage}
  \maketitle
  \end{titlepage}

\begin{abstract}
The mental contexts in which we interpret experiences are often person-specific, even when the experiences themselves are shared. We developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences. We model experiences and memories as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration. Memory encoding can then be modeled as geometrically preserving or distorting the \textit{shape} of the original experience. We applied our approach to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging. Participants’ recountings all preserved coarse spatial properties (essential narrative elements), but not fine spatial scale (low-level) details, of the episode’s trajectory. We also identified networks of brain structures sensitive to these trajectory shapes. Our work provides insights into how our brains preserve and distort our ongoing experiences when we encode them into episodic memories.
\end{abstract}

@@

Here, we sought to examine how the temporal dynamics of a naturalistic experience were later reflected in participants' memories.  We also sought to leverage the above conceptual insights into the distinctions between an experience's essence and its low-level details to build models that explicitly quantified these distinctions.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television show \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Our framework uses topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls by projecting each moment into a word embedding space.  We then use hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast both naturalistic experiences and memories of those experiences as geometric \textit{trajectories} through word embedding space that describe how they evolve over time. Under this framework, successful remembering entails verbally traversing the content trajectory of the episode, thereby reproducing the shape (essence) of the original experience.  Our framework captures the episode's essence in the sequence of geometric coordinates for its events, and its low-level details by examining its within-event geometric properties.

@@

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall transcript, we treated each overlapping  window of (up to) 10 sentences from their transcript as a document, and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  An important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite that different participants used widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This ability to match up conceptually related text that differs in specific vocabulary, detail, and length is an important benefit of projecting the episode and recalls into a shared topic space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions matrix (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

@@

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the episode, and the events they later recalled.  We labeled each recalled event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transitioned between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with early events (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging meaningful narrative associations and references over long timescales throughout the episode.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (P17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of recalled episode events, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of recalled episode events, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision reflects the completeness of recall, whereas distinctiveness reflects recall specificity.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  episode events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} Excerpts from the most precise (P17) and least precise (P6) participants' recalls of episode event 22.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.  \textbf{D.} Recall distinctiveness by episode event.  Kernel density estimates for each episode event's distribution of recall distinctiveness scores, analogous to Panel A.  \textbf{E.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in Panel F.  Each event's text is highlighted in a different color.  \textbf{F.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 22.  Sections of recall describing each each episode event in Panel E are highlighted with the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

Examining individual recalls of the same episode event can provide insights into how the above precision and distinctiveness scores may be used to characterize similarities and differences in how different people describe the same shared experience.  In Figure \ref{fig:precision-detail}, we compare recalls for the same episode event (event 22) from different participants: one with a high precision score (P17), and the other with a low precision score (P6).  From the HMM-identified event boundaries, we recovered the set of annotations describing the content of an example episode event (Fig.~\ref{fig:precision-detail}B), and divided them into different color-coded sections for each action or feature described.  We used an analogous approach to identify the set of sentences comprising the corresponding recall events for each of the two example participants.  Figure~\ref{fig:precision-detail}C shows excerpts of two participants' recall transcripts that comprised sentences between the first and last descriptions of content from the example episode event.  We then colored all words describing actions and features in the transcripts shown in Panel C according to the color-coded annotations in Panel B.  Visual comparison of these example recalls reveals that the more precise recall captures more of the episode event's content, and in greater detail.

Figure \ref{fig:precision-detail} also illustrates the differences between high and low distinctiveness scores for the same event detailed in Figure~\ref{fig:precision-detail}B (i.e., event 22).  Here, we have extracted the set of sentences comprising the most distinctive recall event (P9) and least distinctive recall event (P6) matched to the example episode event (Fig.~\ref{fig:precision-detail}F).  We also extracted the annotations for the example episode event, as well as those from each other episode event whose content the example participants' single recall events described (Fig.~\ref{fig:precision-detail}E).  We assigned each episode event a unique color (Panel E) and colored each recalled phrase or sentence (Panel F) according to the episode events they best matched.  Visual inspection of Panel F reveals that the most distinctive recall's content is tightly concentrated around event 22, whereas the least distinctive recall incorporates content from a much wider range of episode events.

The preceding analyses sought to characterize how participants' recountings of individual episode events captured the low-level details of each event.  Next we sought to characterize how participants' recountings of the full episode captured its high-level essence-- i.e., the shape of the episode's trajectory through word embedding (topic) space.  To visualize the essence of the episode and each participant's recall trajectory~\citep{HeusEtal18a}, we projected the topic proportions matrices for the episode and recalls onto a shared two-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInEtal18}.  In this lower-dimensional space, each point represents a single episode or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}). In other words, events that are nearer to each other in this space are more semantically similar, and those that are farther apart are less so.

@@

In addition to enabling us to visualize the episode's high-level essence, describing the episode as a geometric trajectory also enables us to drill down to individual words and quantify how each word relates to the memorability of each event.  This provides another approach to examining participants' recall for low-level details beyond the precision and distinctiveness measures we defined above.  The results displayed in Figures \ref{fig:list-learning}C and \ref{fig:precision-detail}A suggest that certain events were remembered better than others.  Given this, we next asked asked whether the events were generally remembered precisely or imprecisely tended to reflect particular content.  Because our analysis framework projects the dynamic episode content and participants' recalls into a shared space, and because the dimensions of that space represent topics (which are, in turn, sets of weights over known words in the vocabulary), we are able to recover the weighted combination of words that make up any point (i.e., topic vector) in this space.  We first computed the average precision with which participants recalled each of the 30 episode events (Fig.~\ref{fig:topics}A; note that this result is analogous to a serial position curve created from our precision metric).  We then computed a weighted average of the topic vectors for each episode event, where the weights reflected how precisely each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by more precisely-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Across the full episode, content that weighted heavily on topics and words central to the major foci of the episode (e.g., the names of the two main characters, ``Sherlock'' and ``John,'' and the address of a major recurring location, ``221B Baker Street'') were best remembered.  An analogous analysis revealed which themes were less-precisely remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to its average precision (Fig.~\ref{fig:topics}B, red box).  The least precisely remembered episode content reflected information that was extraneous to the episode's essence, such as the proper names of relatively minor characters (e.g., ``Mike,'' ``Molly,'' and ``Lestrade'') and locations (e.g., ``St. Bartholomew's Hospital'').

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/topics}
\caption{\small \textbf{Language used in the most and least precisely remembered events.} \textbf{A.} Average precision (episode event-recall event topic vector correlation) across participants for each episode event.  Here we defined each episode event's precision for each participant as the correlation between its topic vector and the most-correlated recall event's topic vector from that participant.  Error bars denote bootstrap-derived across-participant 95\% confidence intervals.  The stars denote the three most precisely remembered events (green) and least precisely remembered events (red).  \textbf{B.} Wordles comprising the top 200 highest-weighted words reflected in the weighted-average topic vector across episode events.  Green: episode events were weighted by their precision (Panel A).  Red: episode events were weighted by the inverse of their precision.  \textbf{C.}  The set of all episode and recall events is projected onto the two-dimensional space derived in Figure~\ref{fig:trajectory}.  The dots outlined in black denote episode events (dot size is proportional to each event's average precision).  The dots without black outlines denote individual recall events from each participant.  All dots are colored using the same scheme as Figure~\ref{fig:trajectory}A.  Wordles for several example events are displayed (green: three most precisely remembered events; red: three least precisely remembered events).  Within each circular wordle, the left side displays words associated with the topic vector for the episode event, and the right side displays words associated with the (average) recall event topic vector, across all recall events matched to the given episode event.}
\label{fig:topics}
\end{figure}

A similar result emerged from assessing the topic vectors for individual episode and recall events (Fig.~\ref{fig:topics}C).  Here, for each of the three most and least precisely remembered episode events, we have constructed two wordles: one from the original episode event's topic vector (left) and a second from the average recall topic vector for that event (right).  The three most precisely remembered events (circled in green) correspond to scenes integral to the central plot-line: a mysterious figure spying on John in a phone booth; John meeting Sherlock at Baker St.\~to discuss the murders; and Sherlock laying a trap to catch the killer.  Meanwhile, the least precisely remembered events (circled in red) reflect scenes that comprise minor plot points: a video of singing cartoon characters that participants viewed in an introductory clip prior to the main episode; John asking Molly about Sherlock's habit of over-analyzing people; and Sherlock noticing evidence of Anderson's and Donovan's affair.

The results thus far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories.  We next carried out a series of analyses aimed at understanding which brain structures might facilitate these preservations and transformations between the participants' shared experience of watching the episode and their subsequent memories of the episode.  In the first analysis, we sought to identify brain structures that were sensitive to the dynamic unfolding of the episode's content, as characterized by its topic trajectory.  We used a searchlight procedure to identify clusters of voxels whose activity patterns displayed a proximal temporal correlation structure (as participants watched the episode) matching that of the original episode's topic proportions (Fig.~\ref{fig:brainz}A; see \textit{Methods} for additional details).  In a second analysis, we sought to identify brain structures whose responses (during episode viewing) reflected how each participant would later structure their \textit{recounting} of the episode.  We used a searchlight procedure to identify clusters of voxels whose proximal temporal correlation matrices matched that of the topic proportions matrix for each participant's recall transcript (Figs.~\ref{fig:brainz}B; see \textit{Methods} for additional details).  To ensure our searchlight procedure identified regions \textit{specifically} sensitive to the temporal structure of the episode or recalls (i.e., rather than those with a temporal autocorrelation length similar to that of the episode and recalls), we performed a phase shift-based permutation correction (see \textit{Methods} for additional details). As shown in Figure~\ref{fig:brainz}C, the episode-driven searchlight analysis revealed a distributed network of regions that may play a role in processing information relevant to the narrative structure of the episode.  Similarly, the recall-driven searchlight analysis revealed a second network of regions (Fig.~\ref{fig:brainz}D) that may facilitate a person-specific transformation of one's experience into memory.  The top ten \texttt{Neurosynth} terms~\citep{YarkEtal11} associated with each (unthresholded) map are displayed in each panel.  In identifying regions whose responses to ongoing experiences reflect how those experiences will be remembered later, this latter analysis extends classic \textit{subsequent memory effect analyses}~\citep[e.g.,][]{PallWagn02} to the domain of naturalistic experiences.
